Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-17 draft_16Jan/hiscurrent.tex
Changed abstract and conclusion

Experimenting with biological, physical, and artificial systems to generate a more informative dynamical model is a well-established practice in modern science. Traditional methods for modelling physical systems are based on laws of physics that are based either on empirical relationships or intuition. For systems that evolve with time, physical laws yield mathematical equations that govern how the quantities evolve with time. However, our world is much more complex than that which can be distilled into elegant equations. We do not fully understand many of the more complex systems, nor do they provide us with a good physical intuition of the underlying principles governing system dynamics. To complicate this, the underlying systems we observe often display sensitive dependence on initial conditions despite having highly similar initial conditions. This is because differing orbits diverge quickly and to such an extent that it becomes seemingly impossible to retrace their steps back to the original conditions. Even the presence of usually' negligible' computational noise or measurement error renders long-term pointwise prediction infeasible. There are many difficulties that one encounters while modelling such systems:
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item One may not have access to the complete states of the systems
  \item The system may be described by functions that behave wildly because their graphs have a wild oscillatory behaviour, i.e., a large functional complexity.
\end{enumerate}

Models derived primarily from data can be classified into three categories: 
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item  Interpretable models (i.e., they establish relationships between internal physical mechanisms), 
  \item Partially interpretable models capturing some modes of the dynamics, 
  \item Non-Interpretable models, defined as such mainly due to the fact that they are defined on a different phase space that is usually high-dimensional. 
\end{enumerate}

@@

Recent partially interpretable models available in the literature have been based on the Koopman operator (e.g.,[?, ?]) to employ observables mapping the data onto a higher-dimensional space. This makes the dynamics in the higher-dimensional space more amenable for approximation by a linear transformation. Such methods do not guarantee exact reconstructions for nonlinear models, and in practice, provide poor long-term consistency for a large class of systems. (Citation)
\textbf{(Citation)}

The non-interpretable models include the delay embedding and the machine learning algorithms. For example, one could learn a system conjugate to the underlying system by applying the Takens delay embedding~\cite{takens1981detecting} when one has "good" observations from the system; this learnt system could then be used to forecast the observed data. 
Takens delay embedding theorem ~\cite{takens1981detecting} and its various generalisations (e.g., \cite{sauer1991embedology, stark1999delay, gutman2018embedding}) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then confirms the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality still depends on numerous parameters, making the comprehension of the dynamics unreliable (e.g., [?]). One reason for this fragility is that the embedded attractor in the reconstruction space is not always an attractor of the map learnt in the reconstruction space, despite unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter~ \ref{ch3}), it can cause predictions to fail.

Practically, the application of Takens embedding involves learning a map through some technique, and consequently one wishes that these would have low functional complexity\cite{manjunath2021universal}, i.e., functions with fewer oscillatory graphs. On the other hand, pure machine learning methodology processes temporal information (like the echo state networks (Citations to echo state networks)) by mapping data onto a higher dimensional space for further processing. Although they perform  well on forecasting some dynamical systems, they fail completely on others (Citations here) as there is often no guarantee that the right function was learnt during training.

This thesis involves implementing and analysing non-interpretable models that can guarantee exact reconstruction. 
The project work concerns the study and implementation of a method (\cite{manjunath2021universal}) that incorporates learning a function by mapping the data on to a higher dimensional space using what is called a driven dynamical system (See Chapter~ref{ch.4}). With a clear understanding of how the data is mapped onto the higher dimensional space, the method then permits the learning of a dynamical system topologically conjugate to that of the underlying system. Instead of linear regression as in the training of echo state networks, deep learning methods are employed to learn the correct function. With slight modifications to the implementation in the paper (with Adriaan), we show that one can construct accurate non-interpretable models with the ability to reconstruct attractors from more hard-to-forecast systems like the double pendulum. (The forecasting of the time-series from a double pendulum has not been reported before.) Moreover, we also demonstrate that long-term statistical consistency is preserved.

@@

The more intricate details of proofs are referred to where relevant throughout. This report is organised as follows: 
\newline In Chapter~\ref{ch2}, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE and then proceed to define the inverse-limit space and topological conjugacy of autonomous systems. 
\newline Chapter~\ref{ch3} introduces the problem of forecasting dynamical systems, states the Takens delay embedding theorem, and discusses various issues faced while forecasting. 
\newline In Chapter~\ref{ch4}, a driven dynamical system is defined and discuss the properties of a specific class of driven dynamical systems that we make use of in this thesis/report/paper. 
\newline Finally, in Chapter~\ref{ch5} we show the implementation of these forecasting methods, and conclusions are provided in Chapter~\ref{ch6}.

@@

At its most elementary level, a dynamical system is just something that evolves deterministically through time. In the context of this thesis, deterministic refers to the fact that a system evolves according to specified rules rather than based on random events. 
Dynamical systems arise in a variety of situations. A continuous-time dynamical system describes the states for all values of the time. Specifically, if the motion of a pendulum in which the quantities such as the angular position and angular momentum are known at all times, then it is a continuous-time dynamical system. The equations of the dynamical system can take the form of one or more ordinary differential equations that determine the relevant quantities at any future time if we know the initial location and momentum. 
In ecology, discrete-time dynamical systems are widely used to model population growth. The model in this case is a function calculating the following generation's population given the population of the previous generation. If we know the starting population, we may once again calculate the population at any time in the future.

@@

Of course, discrete-time dynamical systems need not always arise through a differential equation. Once again, we may consider the field of ecology, where discrete dynamical systems are often directly derived or assumed. \textbf{Citation}
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations. We refer the interested reader to \cite{saber2010introduction}.

@@

A dynamical system can have several attractors and may also be contained in another attractor. For the example, $Tu =u^2$ on $[0,1]$, both $\{0\}$ and $[0,1]$ are attractors. It is known that an attractor for a dynamical system always exists in a compact space. \textbf{(Cite).}

The dynamics restricted to an invariant set can be complicated. For instance an invariant set could be just a single point or it could have an infinite set. If the invariant set is infinite, then complicated dynamics are possible.A particular well-studied phenomenon of such complexity gives rise to so-called  chaotic behaviour, studied in detail over the past fifty years.

@@

There are several intricately different definitions of chaos available in the literature \textbf{(Cite)}, with each of them indicating some aspect of complexity.
In practice, it is not possible to verify which definition or notion does the underlying dynamical system satisfies when only data is observed from it. We do, however, specifically recall the definition of chaos in the sense of Devaney \cite{devaney2018introduction,de2013elements} so as to understand some nuances behind the complexity. Devaney's definition of chaos has three requirements, with the first being the notion  of sensitive dependence on initial conditions.

@@

\begin{Example} \rm
  The logistic map(LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$ is of particular interest to us as it is classically used to illustrate the influence of chaos on a system \textbf{(Cite)}. 
  In the graph below~\ref{fig:log_sdic} we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$.

@@

\ednote{B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough?  M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly.}
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way ''dynamically equivalent''. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (e.g., \cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

@@

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold)\ednote{B: Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) i.e.\ednote{two i.e.'s}, we examine  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ represents a probe inserted into a bigger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the current weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

@@

By generic we mean a residual set on a certain topology on appropriate spaces of functions (that we do not describe here) \textbf{(Cite)}. 
By $C^2$, we make reference to a twice-differentiable function  with a continuous second derivative. In our scenario (and this remains important throughout) the input space $U$ is considered the attractor.~\label{attractor_U}

@@

This requirement is not overly restrictive, as a great number of chaotic systems do have a surjective map. (Cite) \ednote{B: I remember you saying this in one of our sessions, but don't have a source for it.}

@@

The reachable set itself can defined independently from $g$ possessing the ESP/USP. We note that $x\in{X_n(\overline{u})}$ if and only $g$ has a solution $\{x_k\}$ for $x_n=x$ and input $\overline{u}$. \textbf{(Cite)}.

@@

It turns out that $\bigcap_{i<n}X_{n,i}(\overline{u})$ is identically equal to $X_n((\overline{u})$ that we had defined earlier. This will lead to a result (for. e.g., \cite{manjunath2013echo}) that  \textit{$g$ being a topological contraction is equivalent to the existence of a exactly one entire solution.}

@@

Despite the ease that with which one manipulates a left-infinite history in the realm of theory, it is impossible to obtain or use such a sequence in any real-life application.  
Fortunately, one does not need the entire left-infinite history of an input in practice thanks to the Uniform Attraction Property(UAP). 
We use an alternate version of the definition of UAP -- since the notion of nonautonomous attractors in this thesis would take some time to establish and detracts from the principal thrust of this project, we refer to \cite{Manju_Nonlinearity}.

@@

\begin{figure}[ht]
  \includegraphics[scale=0.3]{_summarypictorial.eps}
  \centering
  \captionof{figure}{Pictorial summary of results obtained up to now and how they overlap. Reproduced from \textbf{(Cite SIG-25May)} }
  \label{fig:fig_pictorialSummary}
\end{figure}

@@

To circumvent this: if a driven system with a saturation function\ednote{B: What is a saturation function?} is used, we can indirectly (via $g$) ensure the existence of an attractor $A$ of the map which implements $G_T$, and containing $Y_T^+$.  \ednote{B: I'm not sure this sentence makes complete sense.}
By defining the dynamics to $A$, this prevents large errors from occuring.  
More details on the global dissipativity of $g$, as it is termed there, may be found in~\cite{Supp}.

@@

We implement a feedforward network, (a network with no memory, unlike the RNN)\ednote{B: Added so that FFN is in more context} required to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library \ednote{B: Is this something you want me to write in? M: Note! We do not implement the RNN, but the FN that learns Gamma through the Tensorflow. We called the RNN + FN set up a RCN in Adriaan's paper. FYI: FN's work only if there is a definite relationship between the $u_n$ and $u_{n+1}$ which is rare in temporal data.
RNNs with ESP work when there is a definite relationship between $(\ldots,u_{n-1},u_n)$ and $u_{n+1}$ and is suited for temporal processing.}.

@@

The two model equations for learning from data (\eqref{eqns_from_data}) allows us to determine what the state $x_{n+1}$ at time $n+1$ will be. 
We specifically remind ourselves of \[x_{k+1}=\pi_2 \circ (\Gamma, \pi_2) \circ (\pi_2,g) (u_k,x_k)\]

@@

We also consider a more general problem of learning an attractor of dynamical system when only observations of an orbit are given. Explicitly, if $(W,T)$ is a dynamical system with dynamics generated by $w_{n+1}=Tw_n$, and if $\theta:W \to \mathbb{R}$ is an observable, then the task is to learn a dynamical system that is topologically conjugate to $(W,T)$ and predict $\theta(w_{m+1}),\theta(w_{m+2}),...$ using the data $\theta(w_{0}),\theta(w_{1}),\ldots,\theta(w_{m})$.  So how do we do it?
Suppose the input generated from the delay-coordinate map $\Phi_{\theta,2d}(\theta(w_{n})) := (\theta(w_{n-2d}),\ldots,\theta(w_{n-1}),\theta(w_{n}))$ is such that Takens delay embedding theorem (see Theorem~\ref{Thm_Takens}) holds, in which case, there exists a homeomorphism  $F_\theta: \Phi_{\theta,2d}(\theta(w_{n})) \mapsto \Phi_{\theta,2d}(\theta(w_{n+1}))$. Then if we feed  the input values $u_n := \Phi_{\theta,2d}(\theta(w_{n}))$ to a driven system $g$ that is SI-invertible and has USP, the induced system $(Y_F,G_F)$ would be topologically conjugate to the inverse-limit space of
$(\Phi_{\theta,2d}(W), F_\theta)$ as in Theorem~\ref{Thm_CET} \ednote{M: please see if this refers to the causal embedding theorem in the thesis}, and hence one could forecast $u_n,u_{n+1},\ldots$. In other words, one could forecast  the values $\theta(w_n), \theta(w_{n+1}),\ldots$ by feeding these observations as delay coordinates. The advantage of feeding delay-coordinates to a driven system is that the embedding is stable in the sense of global dissipativity that we have described in the previous chapter, i.e., $Y_F$ is made to contained in an attractor of a mapping induced on $Y_F$ that is induced by $g$.

@@

Our implementation of the reservoir(A) and input(B) matrices in the RNN makes use of sparse matrices. We discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opted to enforce the requirement that both matrices be well-conditioned, i.e. that they have a condition number of \~1. The condition number of a matrix $M$ (denoted $cond(M)$) is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed. \textbf{(Cite.)} By ensuring a condition number close to 1, we also ensure that our driven function $g$, depending on the matrices $A$ and $B$, will not be sensitive to adjustments in the input. In so doing we then also guarantee a measure of stability. \ednote{B: Not satisfactory. Edit again.}

Lastly, we ensure the matrix $B$ has unit spectral radius. This is necessary due to the Causal Embedding Theorem~\ref{Thm_GT_Exists} which can then guarantee the existence of semi-conjugacy between the systems $(Y_T, G_T)$ and $(\widehat{U}_T, \widehat{T})$. In fact, the theorem actually requires that $\alpha{B}$ have spectral radius $<1$, so we impose the restriction $\alpha<1$ and consider only the matrix $B$'s spectral radius from thereon.

Note that to learn $\Gamma: (x_{n-1},x_{n}) \mapsto u_n$ throughout feedforward network, one could learn from the principal components of $(x_{n-1},x_{n})$ as well. 
We employ this since empirically it gives slightly better results and not essential.
The principal components are used only for an efficient state representation possibly to reduce the errors while learning and is not for a lossy approximation since all principal components are used.
  To be explicit, we denote $X_{1:N}$ to be the matrix with the first $N$ states of the network data as row vectors. So, if $X_{1:N}=U\Sigma P^T$ represents the singular value decomposition of $X_{1:N}$, then the principal component matrix is denoted by $P$, and the principal components are given by 
 $Z_{1:N}=X_{1:N}P.$
These principal components are then used to train a feedforward neural network. If we denote the row vectors of $Z_{1:N}$ by 
$z_i^T, i=1,2,\cdots,N$ and the neural network by $NN$, then the network is trained to learn an approximation of the map
$NN: (z_{n-1},z_n) \mapsto  u_n.
$
The learnt approximation of $NN$ can be used to approximate $\Gamma$ since 
$$
NN\left( \begin{bmatrix} 
P^Tx_{n-1} \\
P^Tx_n
\end{bmatrix}
\right) = NN \circ 
\begin{bmatrix}
P^T & 0 \\
0 & P^T 
\end{bmatrix}\begin{bmatrix}
x_{n-1}\\
x_n
\end{bmatrix} = u_n = \Gamma(x_{n-1},x_n).
$$

@@

\chapter{Conclusions} \label{ch6}

>>>>

Experimenting with biological, physical, and artificial systems to generate a more informative dynamical model is a well-established practice in modern science. Traditional methods for modelling physical systems are based on laws of physics that are based either on empirical relationships or intuition. For systems that evolve with time, physical laws yield mathematical equations that govern how the quantities evolve with time. However, our world is much more complex than that which can be distilled into elegant equations. We do not fully understand many of the more complex systems, nor do they provide us with a good physical intuition of the underlying principles governing system dynamics. To complicate this, the underlying systems we observe often display sensitive dependence on initial conditions despite having highly similar initial conditions. This is because differing orbits diverge quickly and to such an extent that it becomes seemingly impossible to retrace their steps back to the original conditions. Even the presence of usually `negligible' computational noise or measurement error renders long-term pointwise prediction infeasible. We encounter many difficultiesin the process of modelling such systems:
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item One may not have access to the complete states of the systems
  \item The system may be described by functions that behave wildly because their graphs have a wild oscillatory behaviour, i.e., a large functional complexity.
\end{enumerate}

Models derived primarily from data can be classified into three categories: 
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item  Interpretable models (i.e., they establish relationships between internal physical mechanisms), 
  \item Partially interpretable models capturing some modes of the dynamics, 
  \item Non-interpretable models, defined as such mainly due to the fact that they are defined on a different phase space that is usually high-dimensional. 
\end{enumerate}

@@

Recent partially interpretable models available in the literature have been based on the Koopman operator (see~\cite{koopman1932dynamical,budivsic2012applied}) to employ observables mapping the data onto a higher-dimensional space. This makes the dynamics in the higher-dimensional space more amenable for approximation by a linear transformation. Such methods do not guarantee exact reconstructions for nonlinear models, and in practice provide poor long-term consistency for a large class of chaotic dynamical systems \cite{Supp}.

The non-interpretable models include the delay embedding and the machine learning algorithms. For example, one could learn a system conjugate to the underlying system by applying the Takens delay embedding method of delay-coordinates~\cite{takens1981detecting} when one has "good" observations from the system; this learnt system could then be used to forecast the observed data. 
Takens delay embedding theorem~\cite{takens1981detecting} and its various generalisations (see,~\cite{sauer1991embedology, stark1999delay, gutman2018embedding}) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then confirms the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality still depends on numerous parameters, making the comprehension of the dynamics unreliable (see, [?]). One reason for this fragility is that the embedded attractor in the reconstruction space is not always an attractor of the map learnt in the reconstruction space, despite unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter~ \ref{ch3}), it can cause predictions to fail.

Practically, the application of Takens embedding involves learning a map through some technique, and consequently one wishes that these would have low functional complexity\cite{manjunath2021universal}, i.e., functions with fewer oscillatory graphs. On the other hand, pure machine learning methodology processes temporal information (like the echo state networks~\cite{Manju_ESP, Manju_IEEE, grigoryeva2018echo}) by mapping data onto a higher dimensional space for further processing. Although they perform  well on forecasting some dynamical systems, they fail completely on others (\ednote{Ask for resources here}) as there is often no guarantee that the right function was learnt during training.

This project involves implementing and analysing non-interpretable models that can guarantee exact reconstruction. 
The project work concerns the study and implementation of a method (\cite{manjunath2021universal}) that incorporates learning a function by mapping the data on to a higher dimensional space using what is called a driven dynamical system (See Chapter~ref{ch.4}). With a clear understanding of how the data is mapped onto the higher dimensional space, the method then permits the learning of a dynamical system topologically conjugate to that of the underlying system. Instead of linear regression as in the training of echo state networks, deep learning methods are employed to learn the correct function. With slight modifications to the implementation in the paper (\cite{manjunath2021universal}), we show that one can construct accurate non-interpretable models with the ability to reconstruct attractors from more hard-to-forecast systems like the double pendulum. (The forecasting of the time-series from a double pendulum has not been reported before.) Moreover, we also demonstrate that long-term statistical consistency is preserved.

@@

The more intricate details of proofs are referred to where relevant throughout. This report is organised as follows: 
\newline In Chapter~\ref{ch2}, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE and then proceed to define the inverse-limit space and topological conjugacy of autonomous systems. 
\newline Chapter~\ref{ch3} introduces the problem of forecasting dynamical systems, states the Takens delay embedding theorem, and discusses various issues faced while forecasting. 
\newline In Chapter~\ref{ch4}, a driven dynamical system is defined and discuss the properties of a specific class of driven dynamical systems that we make use of in this project.
\newline Finally, in Chapter~\ref{ch5} we show the implementation of these forecasting methods, and conclusions are provided in Chapter~\ref{ch6}.

@@

At its most elementary level, a dynamical system is just something that evolves deterministically through time. In the context of this project, deterministic refers to the fact that a system evolves according to specified rules rather than based on random events. 
Dynamical systems arise in a variety of situations. A continuous-time dynamical system describes the states for all values of the time. Specifically, if the motion of a pendulum in which the quantities such as the angular position and angular momentum are known at all times, then it is a continuous-time dynamical system. The equations of the dynamical system can take the form of one or more ordinary differential equations that determine the relevant quantities at any future time if we know the initial location and momentum. 
In ecology, discrete-time dynamical systems are widely used to model population growth. The model in this case is a function calculating the following generation's population given the population of the previous generation. If we know the starting population, we may once again calculate the population at any time in the future.

@@

Of course, discrete-time dynamical systems need not always arise through a differential equation. Once again, we may consider the field of ecology, where discrete dynamical systems are often directly derived or assumed. \ednote{Ask for resource}
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations. We refer the interested reader to~\cite{saber2010introduction}.

@@

A dynamical system can have several attractors and may also be contained in another attractor. For the example, $Tu =u^2$ on $[0,1]$, both $\{0\}$ and $[0,1]$ are attractors. It is known that an attractor for a dynamical system always exists in a compact space~\cite{Milnor1985}.

The dynamics restricted to an invariant set can be complicated. For instance an invariant set could be just a single point or it could have an infinite set. If the invariant set is infinite, then complicated dynamics are possible. A particular, well-studied phenomenon of such complexity gives rise to so-called chaotic behaviour, a subject studied in detail over the past fifty years.

@@

Different authors propose a number of intricately different definitions of chaos in the literature~\cite{RasbandChaos, TaborChaos, WigginsChaos} and each of them indicate some aspect of complexity.

In practice, when only data is observed from a system, it is not possible to verify which definition or notion is satisfied by the underlying dynamical system. We do, however, specifically recall the definition of chaos in the sense of Devaney \cite{devaney2018introduction,de2013elements} so as to understand some nuances behind the complexity. Devaney's definition of chaos has three requirements, with the first being the notion  of sensitive dependence on initial conditions.

@@

\begin{Example} \rm
  The logistic map(LM), a recursion relation of the form $x_{n+1}=rx_n(1-x_n)$ where $x_n\in[0,1]$ and $r\leq{4}$ is a classic example used to illustrate the chaotic behaviour that a system may exhibit. When $r=4$, the one-dimensional recurrence relation $x_{n+1}=4x_n(1-x_n)$ can be used as the kindergarten's model to exhibit the presence of SDIC. \ednote{B: edited and removed unnecessary citation}
  In the graph below~\ref{fig:log_sdic} we plot the first 50 iterates of the LM for $r=4$ and two slightly different initial values $x_0$.

@@

\ednote{B: Does defining explicitly the domain and codomain of $\phi$ make it clear enough?  M: Saying $\phi$ is then a semi-conjugacy between $T$ and $S$ could be ambiguous as it does not specify which system is on the top in the commutativity. So please edit accordingly.}
When $S$ is conjugate to $T$, the dynamics of the two systems are in some way `dynamically equivalent'. Specifically, they are in one-to-one correspondence with one another. However when $S$ is semi-conjugate to $T$ with $\phi:U\to{V}$ a many-to-one mapping, the dynamics on $V$ provide merely a coarse-grained description of the dynamics on $U$~\cite{de2013elements}. When $S$ is semi-conjugate to $T$, it is also common to call $S$ a \emph{factor of $T$}, or conversely that $T$ is an \emph{extension of $S$}. In essence, an extension (see, \cite{de2013elements}) is a more extensive system capturing all of the essential dynamics of its factor.

@@

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold)\ednote{B: Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) by examining  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ represents a probe inserted into a bigger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the current weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

@@

By generic we mean a residual set on a certain topology on appropriate spaces of functions (that we do not describe here) \ednote{B: Where can I find a citation for this?}. 
By $C^2$, we make reference to a twice-differentiable function  with a continuous second derivative. In our scenario (and this remains important throughout) the input space $U$ is considered the attractor.~\label{attractor_U}

@@

This requirement is not overly restrictive, as a great number of chaotic systems do have a surjective map. (Cite) \ednote{B: I remember you saying this in one of our sessions, but don't have a source for it. Do you have a specific reference?}

@@

The reachable set itself can defined independently from $g$ possessing the ESP/USP. Note that $x\in{X_n(\overline{u})}$ if and only $g$ has a solution $\{x_k\}$ for $x_n=x$ and input $\overline{u}$, a result proved in nonautonomous dynamical systems literature~\cite{manjunath2014dynamics, manjunath2013echo}. \ednote{B: Adjusted sentence and cited}

@@

It turns out that $\bigcap_{i<n}X_{n,i}(\overline{u})$ is identically equal to $X_n((\overline{u})$ that we had defined earlier. This will lead to a result (for. see, \cite{manjunath2013echo}) that  \textit{$g$ being a topological contraction is equivalent to the existence of a exactly one entire solution.}

@@

Despite the ease that with which one manipulates a left-infinite history in the realm of theory, it is impossible to obtain or use such a sequence in any real-life application.  
Fortunately, one does not need the entire left-infinite history of an input in practice thanks to the Uniform Attraction Property(UAP). 
We use an alternate version of the definition of UAP -- since the notion of nonautonomous attractors in this project would take some time to establish and detracts from the principal thrust of this project, we refer to \cite{Manju_Nonlinearity}.

@@

\begin{figure}[ht]
  \includegraphics[scale=0.3]{_summarypictorial.eps}
  \centering
  \captionof{figure}{Pictorial summary of results obtained up to now and how they overlap. Reproduced from \textbf{(Cite SIG-25May)} }
  \label{fig:fig_pictorialSummary}
\end{figure}
\ednote{Ask M: Cite SIG-25May.}

@@

To circumvent this: if a driven system with a saturation function\ednote{B: What is a saturation function?} is used, we can ensure the existence of an attractor $A$ of the map which implements $G_T$, and containing $Y_T^+$ indirectly (via $g$).  \ednote{B: A little vague. Are we saying that an attractor of a map implementing $G_T$ is also the one that contains $Y_T^+$?}
By defining the dynamics to $A$, this prevents large errors from occuring.  
More details on the global dissipativity of $g$, as it is termed there, may be found in~\cite{Supp}.

@@

We implement a feedforward network, (a network with no memory, unlike the RNN)\ednote{B: Expanded sentence s.t. FFN is in more context} required to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library \ednote{B: Is this something you want me to write in? M: We do not implement the RNN, but the FN that learns Gamma through the Tensorflow. We called the RNN + FN set up a RCN in Adriaan's paper. FYI: FN's work only if there is a definite relationship between the $u_n$ and $u_{n+1}$ which is rare in temporal data.
RNNs with ESP work when there is a definite relationship between $(\ldots,u_{n-1},u_n)$ and $u_{n+1}$ and is suited for temporal processing.}.

@@

The two model equations for learning from data~\eqref{eqns_from_data} allows us to determine what the state $x_{n+1}$ at time $n+1$ will be. 
We specifically remind ourselves of \[x_{k+1}=\pi_2 \circ (\Gamma, \pi_2) \circ (\pi_2,g) (u_k,x_k)\]

@@

We also consider a more general problem of learning an attractor of dynamical system when only observations of an orbit are given. Explicitly, if $(W,T)$ is a dynamical system with dynamics generated by $w_{n+1}=Tw_n$, and if $\theta:W \to \mathbb{R}$ is an observable, then the task is to learn a dynamical system that is topologically conjugate to $(W,T)$ and predict $\theta(w_{m+1}),\theta(w_{m+2}),...$ using the data $\theta(w_{0}),\theta(w_{1}),\ldots,\theta(w_{m})$.  So how do we do it?
Suppose the input generated from the delay-coordinate map $\Phi_{\theta,2d}(\theta(w_{n})) := (\theta(w_{n-2d}),\ldots,\theta(w_{n-1}),\theta(w_{n}))$ is such that Takens delay embedding theorem (see Theorem~\ref{Thm_Takens}) holds, in which case, there exists a homeomorphism  $F_\theta: \Phi_{\theta,2d}(\theta(w_{n})) \mapsto \Phi_{\theta,2d}(\theta(w_{n+1}))$. Then if we feed  the input values $u_n := \Phi_{\theta,2d}(\theta(w_{n}))$ to a driven system $g$ that is SI-invertible and has USP, the induced system $(Y_F,G_F)$ would be topologically conjugate to the inverse-limit space of
$(\Phi_{\theta,2d}(W), F_\theta)$ as in Theorem~\ref{Thm_CET}, and hence one could forecast $u_n,u_{n+1},\ldots$. In other words, one could forecast  the values $\theta(w_n), \theta(w_{n+1}),\ldots$ by feeding these observations as delay coordinates. The advantage of feeding delay-coordinates to a driven system is that the embedding is stable in the sense of global dissipativity that we have described in the previous chapter, i.e., $Y_F$ is made to contained in an attractor of a mapping induced on $Y_F$ that is induced by $g$.

@@

Our implementation of the reservoir(A) and input(B) matrices in the RNN makes use of sparse matrices. We discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opted to enforce the requirement that both matrices be well-conditioned, i.e. that they have a condition number of \~1. The condition number of a matrix $M$ (denoted $cond(M)$) is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed. \textbf{(Cite.)}\ednote{Ask M: Cite specific resource?} By ensuring a condition number close to 1, we also ensure that our driven function $g$, depending on the matrices $A$ and $B$, will not be sensitive to adjustments in the input. In so doing we then also guarantee another form stability. \ednote{B: Not complete enough. Edit further and cite}

Lastly, we ensure the matrix $B$ has unit spectral radius. This is necessary due to the Causal Embedding Theorem~\ref{Thm_CET} which can then guarantee the existence of semi-conjugacy between the systems $(Y_T, G_T)$ and $(\widehat{U}_T, \widehat{T})$. In fact, the theorem actually requires that $\alpha{B}$ have spectral radius $<1$, so we impose the restriction $\alpha<1$ and consider only the matrix $B$'s spectral radius from thereon.

Note that to learn $\Gamma: (x_{n-1},x_{n}) \mapsto u_n$ through our feedforward network, we could learn the principal components of $(x_{n-1},x_{n})$ as well. 
We employ this since it provides slightly better results empirically but is not essential.

Principal components are used only for a more efficient state representation to possibly reduce learning errors; it is not for a lossy approximation since we use all principal components \textbf{(Cite.)}\ednote{Ask M: Is there a specific resource to cite?}.
More explicitly, denote $X_{1:N}$ as the matrix with the first $N$ states of the network data represented by row vectors. If $X_{1:N}=U\Sigma P^T$\ednote{Expand here} represents the singular value decomposition of $X_{1:N}$ \ednote{Ask: Is there a specific resource to cite?}, we then denote the principal component matrix $P$, and the principal components are given by 
 $Z_{1:N}=X_{1:N}P.$
These principal components are then used to train a feedforward neural network. Denote the row vectors of $Z_{1:N}$ by 
$z_i^T, i=1,2,\cdots,N$ and the neural network by $NN$, the network is subsequently trained to learn an approximation of the map
$NN: (z_{n-1},z_n) \mapsto  u_n.
$
The learnt approximation of $NN$ can be used to approximate $\Gamma$ since 
$$
NN\left( \begin{bmatrix} 
P^Tx_{n-1} \\
P^Tx_n
\end{bmatrix}
\right) = NN \circ 
\begin{bmatrix}
P^T & 0 \\
0 & P^T 
\end{bmatrix}\begin{bmatrix}
x_{n-1}\\
x_n
\end{bmatrix} = u_n = \Gamma(x_{n-1},x_n).
$$

@@

\chapter{Conclusion}\label{ch6}
\input{chapter6.tex}
